---
title: "Metadata"
author: "Lisa"
output:
rmarkdown::html_document:
   theme: united
   highlight: tango
   code_folding: hide
   toc: true
   toc_float: true
   df_print: paged
   smooth_scroll: true
   number_sections: false
   self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
setwd("/Volumes/Samsung_T5/MiG_data/MDD_microglia_analysis/MDD_microglia_analysis/AD_analyses")
library(ProjectTemplate)
load.project()
```

```{r packages, echo=FALSE, message=FALSE, results='hide'}
library(UpSetR)
library(readxl)
library(knitr)
library(kableExtra)
library(ggplot2)
library(RColorBrewer)
library(tidyverse)
```

# Preparing metadata 


### Change to correct data format
```{r changing format, echo=TRUE}
metadata <- X202104021_metadata_tidy.Blad1
metadata$Diagnosis <- as.factor(metadata$Diagnosis)
metadata$Main_diagnosis <- as.factor(metadata$Main_diagnosis)
metadata$Region <- as.factor(metadata$Tissue)
#repeat this for AD columns?
```

### Number of unique donors
```{r unique donors, echo=TRUE}
#set rownames to Donor_tissue
# row.names(metadata) <- metadata$Donor_tissue
length(unique(metadata$Donor_id)) 
```

### Number of Antidepressant users 

```{r antidressant users, echo=TRUE}
#create new dataframe with only AD users
metadata_AD <- metadata[metadata[,"Antidepressant_90days"]== "YES",] 
length(unique(metadata_AD$Donor_id))
```
### Number of Antipsychotics users 
```{r}
metadata_AP <- metadata[metadata[,"Antipsychotica_90days"]== "YES",] 
length(unique(metadata_AP$Donor_id))
```



### Number of Patients using both Antidepressants and Antipsychotics 
```{r, echo=TRUE}
metadata_ADAP <- metadata_AD[metadata_AD[, "Antipsychotica_90days"]== "YES",]
length(unique(metadata_ADAP$Donor_id))
donor_AD_AP <- unique(metadata_ADAP$Donor_id)
```

### Number of patients per antidepressant type 
```{r, echo=TRUE}
donor_AD = unique(metadata_AD[,c("Donor_id", "Antidepressant_1", "Antidepressant_2")])

#check which donors only use 1 AD
donor_AD[donor_AD[,"Antidepressant_2"]=='NO',]

as.data.frame(t(as.matrix(unclass(table(c(donor_AD$Antidepressant_1,donor_AD$Antidepressant_2), useNA = "ifany")  )))) %>%
kable(row.names = F) %>%
kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

```

### Number of samples per antidepressant type 
```{r, echo=TRUE}
as.data.frame(t(as.matrix(unclass(table(metadata$Antidepressant_90days, useNA = "ifany")  )))) %>%
kable(row.names = F) %>%
kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

```

```{r, echo=TRUE}
donor_1AD = unique(metadata[,c("Donor_id", "Antidepressant_1")])
pie_table = table(donor_1AD$Antidepressant_1)
myPalette <- brewer.pal(10, "Set3") 
pie(pie_table, border = "white", col = myPalette) +
  coord_polar("y", start = 0) 
```



```{r, results='hide'}
# Main diagnosis by donor 
Main_diagnosisByDonor = unique(metadata[,c("Donor_id", "Main_diagnosis")])
Diagnosisbydonor = unique(metadata[,c("Donor_id", "Diagnosis")])
```

### Number of different main diagnoses 
```{r, echo=TRUE}
length(unique(metadata$Main_diagnosis))
length(unique(metadata$Diagnosis))
```

### Pie chart main diagnosis 
```{r, echo=TRUE}
pie_table = table(Main_diagnosisByDonor$Main_diagnosis)
myPalette <- brewer.pal(10, "Set3") 
pie(pie_table, border = "white", col = myPalette) +
  coord_polar("y", start = 0) 

as.data.frame(t(as.matrix(unclass(table(Main_diagnosisByDonor$Main_diagnosis, useNA = "ifany")  )))) %>%
kable(row.names = F) %>%
kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))



```

```{r}

```




```{r, echo = T}
as.data.frame(t(as.matrix(unclass(table(Diagnosisbydonor$Diagnosis, useNA = "ifany")  )))) %>%
kable(row.names = F) %>%
kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

pie_table = table(Diagnosisbydonor$Diagnosis)
myPalette <- brewer.pal(10, "Set3") 
pie(pie_table, border = "white", col = myPalette) +
  coord_polar("y", start = 0) 
```




```{r, results='hide'}
## table overview of antidepressant types 
table(metadata_AD$Donor_id, metadata_AD$Antidepressant_1)
```

### create dataframe with only TCA users and vector with the donors using TCA's 
```{r, echo=TRUE}
 TCA <- metadata_AD %>% 
   filter(Antidepressant_1 == "TCA")

TCA_donorid <- TCA$Donor_id
TCA_samples <- TCA$Donor_tissue
```

###repeat this for other types 
```{r, echo=TRUE}
 SSRI <- metadata_AD %>% 
   filter(Antidepressant_1 == "SSRI")

SSRI_donorid <- SSRI$Donor_id
SSRI_samples <- SSRI$Donor_tissue

SNRI <- metadata_AD %>% 
   filter(Antidepressant_1 == "SNRI")

SNRI_donorid <- SNRI$Donor_id
SNRI_samples <- SNRI$Donor_tissue

MAOI <- metadata_AD %>% 
   filter(Antidepressant_1 == "MAOI")

MAOI_donorid <- MAOI$Donor_id
MAOI_samples <- MAOI$Donor_tissue

Other <- metadata_AD %>% 
   filter(Antidepressant_1 == "Other")

Other_donorid <- Other$Donor_id
Other_samples <- Other$Donor_tissue

Other2 <- metadata_AD %>% 
   filter(Antidepressant_2 == "Other")

Other_donorid2 <- Other2$Donor_id
Other_samples2 <- Other2$Donor_tissue
```



### Upset plot of AD users by donor 
```{r, echo=TRUE}
list_ADdonor<- list(TCA=TCA_donorid, SSRI=SSRI_donorid, SNRI=SNRI_donorid, MAOI=MAOI_donorid, Other=c(Other_donorid,Other_donorid2))

upset(fromList(list_ADdonor), order.by = "freq")
```

### Upset plot of AD users by sample 
```{r, echo=TRUE}
list_ADsample <- list(TCA=TCA_samples, SSRI=SSRI_samples, SNRI=SNRI_samples, MAOI=MAOI_samples, Other=c(Other_samples,Other_samples2))

upset(fromList(list_ADsample), order.by = "freq")
```

### create dataframes for antipsychotica users as well as vectors 
```{r, echo=TRUE}
Typical <- metadata %>% 
   filter(Antipsychotica_typical == "YES")

Typical_donorid <- Typical$Donor_id
Typical_samples <- Typical$Donor_tissue

Atypical <- metadata %>% 
   filter(Antipsychotica_atypical == "YES")

Atypical_donorid <- Atypical$Donor_id
Atypical_samples <- Atypical$Donor_tissue
```

### Upset plot of donors using AD and/or AP
```{r, echo=TRUE}
list_ADdonor<- list(TCA=TCA_donorid, SSRI=SSRI_donorid, SNRI=SNRI_donorid, MAOI=MAOI_donorid, Other=c(Other_donorid,Other_donorid2), Atypical=Atypical_donorid, Typical=Typical_donorid)

upset(fromList(list_ADdonor), order.by = "freq", nintersects = NA)
```

### Upset plot of samples using AD and/or AP
```{r, echo=TRUE}
list_ADAPsample <- list(TCA=TCA_samples, SSRI=SSRI_samples, SNRI=SNRI_samples, MAOI=MAOI_samples, Other=c(Other_samples,Other_samples2), Atypical=Atypical_samples, Typical=Typical_samples)

upset(fromList(list_ADAPsample), order.by = "freq")
```

### Number of tissue samples in AD group 
```{r, echo =TRUE}
as.data.frame(t(as.matrix(unclass(table(metadata_AD$Tissue, useNA = "ifany")  )))) %>%
kable(row.names = F) %>%
kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

### overview of all samples by main diagnosis
```{r, echo=TRUE}
as.data.frame(table(metadata[,c("Main_diagnosis", "Tissue")]), useNA = "ifany") %>% tidyr::spread(Tissue, Freq) %>%
 kable(row.names = F) %>%
 kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

# Age 
### Maximum minimum and mean age > all donors  
```{r, echo=TRUE}
ageByDonor = unique(ageADorAP[,c("Donor_id", "age", "sex", "Antidepressant_90days", "Antipsychotica_90days")])
#createDT(ageByDonor)
max(ageByDonor$age, na.rm = T)

min(ageByDonor$age, na.rm = T)

mean(ageByDonor$age, na.rm = T)
```
### Maximum minimum and mean age > AD users  
```{r, echo=TRUE}
ageAD = unique(metadata_AD[,c("Donor_id", "age", "sex", "Antidepressant_90days", "Antipsychotica_90days")])

max(ageAD$age, na.rm = T)

min(ageAD$age, na.rm = T)

mean(ageAD$age, na.rm = T)
```
### Maximum minimum and mean age > AP users 
```{r, echo=TRUE}
ageAP = unique(metadata_AP[,c("Donor_id", "age", "sex", "Antidepressant_90days", "Antipsychotica_90days")])

max(ageAP$age, na.rm = T)

min(ageAP$age, na.rm = T)

mean(ageAP$age, na.rm = T)
```
Maximum minimum and mean age > no medication 
```{r, echo=T}
ageNoMed <- ageByDonor[ageByDonor[,"Antidepressant_90days"] == "NO",]
ageNoMed <- ageNoMed[ageNoMed[, "Antipsychotica_90days"]== "NO",]

max(ageNoMed$age, na.rm = T)

mean(ageNoMed$age, na.rm = T)

min(ageNoMed$age, na.rm = T)
```



### Distribution plot age/sex by donor
```{r}
mean_f = mean(ageByDonor[ageByDonor$sex == "f", "age"], na.rm = T)
mean_m = mean(ageByDonor[ageByDonor$sex == "m", "age"], na.rm = T)
ggplot(ageByDonor, aes(x=age, fill=sex)) +
  geom_histogram(bins = 25, colour='black', position = "stack") +
  labs(x="Age", y="Donors") +
  scale_y_continuous(breaks = (1:20)) +
  scale_x_continuous(breaks=seq(20,120,10)) + 
  geom_vline(xintercept=mean_f, color = "red", linetype="dashed") +
  geom_vline(xintercept=mean_m, color = "blue", linetype="dashed") +
  theme_classic()


```

### Distribution plot age/sex by donor using AD 
```{r, echo=TRUE}
mean_f = mean(ageAD[ageAD$sex == "f", "age"], na.rm = T)
mean_m = mean(ageAD[ageAD$sex == "m", "age"], na.rm = T)
ggplot(ageAD, aes(x=age, fill=sex)) +
  geom_histogram(bins = 25, colour='black', position = "stack") +
  labs(x="Age", y="Donors") +
  scale_y_continuous(breaks = (1:20)) +
  scale_x_continuous(breaks=seq(20,120,10)) + 
  geom_vline(xintercept=mean_f, color = "red", linetype="dashed") +
  geom_vline(xintercept=mean_m, color = "blue", linetype="dashed") +
  theme_classic()
```

### Distribution plot age/sex by donor using AP
```{r, echo=TRUE}
mean_f = mean(ageAP[ageAD$sex == "f", "age"], na.rm = T)
mean_m = mean(ageAP[ageAD$sex == "m", "age"], na.rm = T)
ggplot(ageAP, aes(x=age, fill=sex)) +
  geom_histogram(bins = 25, colour='black', position = "stack") +
  labs(x="Age", y="Donors") +
  scale_y_continuous(breaks = (1:20)) +
  scale_x_continuous(breaks=seq(20,120,10)) + 
  geom_vline(xintercept=mean_f, color = "red", linetype="dashed") +
  geom_vline(xintercept=mean_m, color = "blue", linetype="dashed") +
  theme_classic()

```

```{r, previous plot different colors, echo=FALSE}
ggplot(ageAP, aes(x=age, fill=sex)) +
    geom_histogram(bins = 25, colour='black', position = "stack") +
    labs(x="Age", y="Donors") +
    scale_y_continuous(breaks = (1:20)) +
    scale_x_continuous(breaks=seq(20,120,10)) + 
    geom_vline(xintercept=mean_f, color = "red", linetype="dashed") +
    geom_vline(xintercept=mean_m, color = "#9ecae1", linetype="dashed") +
    theme_classic() +
    scale_fill_brewer("sex", c("f", "m"), palette = 1)
```

### Distribution plot age/sex by donor no medication 
```{r, }
mean_f = mean(ageNoMed[ageNoMed$sex == "f", "age"], na.rm = T)
mean_m = mean(ageNoMed[ageNoMed$sex == "m", "age"], na.rm = T)
ggplot(ageNoMed, aes(x=age, fill=sex)) +
  geom_histogram(bins = 25, colour='black', position = "stack") +
  labs(x="Age", y="Donors") +
  scale_y_continuous(breaks = (1:20)) +
  scale_x_continuous(breaks=seq(20,120,10)) + 
  geom_vline(xintercept=mean_f, color = "red", linetype="dashed") +
  geom_vline(xintercept=mean_m, color = "blue", linetype="dashed") +
  theme_classic()
```

### Distribution plot age/sex by donor AD or AP medication 
```{r}
mean_f = mean(ageADorAP[ageADorAP$sex == "f", "age"], na.rm = T)
mean_m = mean(ageADorAP[ageADorAP$sex == "m", "age"], na.rm = T)
ggplot(ageADorAP, aes(x=age, fill=sex)) +
  geom_histogram(bins = 25, colour='black', position = "stack") +
  labs(x="Age", y="Donors") +
  scale_y_continuous(breaks = (1:20)) +
  scale_x_continuous(breaks=seq(20,120,10)) + 
  geom_vline(xintercept=mean_f, color = "red", linetype="dashed") +
  geom_vline(xintercept=mean_m, color = "blue", linetype="dashed") +
  theme_classic()

```


### Number of Euthanasia 
```{r, echo=TRUE}
Euthanasia_donor <- metadata %>% 
   filter(Cause_of_death_categories == "Euthanasia")
#number of donors 
length(unique(Euthanasia_donor$Donor_id))
#number of samples 
length(Euthanasia_donor$Donor_id)
```

###Number of donors with Antidepressiva and Euthanasia
```{r, echo=TRUE}
as.data.frame(t(as.matrix(unclass(table(Euthanasia_donor$Antidepressant_90days, useNA = "ifany")  )))) %>%
kable(row.names = F) %>%
kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

```
###Number of donors with Antipsychotica and Euthanasia 
```{r, echo=TRUE}
as.data.frame(t(as.matrix(unclass(table(Euthanasia_donor$Antipsychotica_90days, useNA = "ifany")  )))) %>%
kable(row.names = F) %>%
kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

```

###Euthanasia with both AP and AD 
```{r, echo=TRUE}
Euthanasia_ADAP <- metadata_ADAP %>% 
   filter(Cause_of_death_categories == "Euthanasia")
#number of donors 
length(unique(Euthanasia_ADAP$Donor_id))
#number of samples 
length(Euthanasia_ADAP$Donor_id)


Euthanasia_ADorAP <- Euthanasia_donor %>% 
   filter(Antipsychotica_atypical == "YES" | Antidepressant_90days == "YES")
length(unique(Euthanasia_ADorAP$Donor_id))

```

###Euthanasia with AP or AD 
```{r, echo=TRUE}
Euthanasia_ADorAP <- Euthanasia_donor %>% 
   filter(Antipsychotica_atypical == "YES" | Antidepressant_90days == "YES")
#number of donors 
length(unique(Euthanasia_ADorAP$Donor_id))
#number of samples 
length(Euthanasia_ADorAP$Donor_id)

```

```{r donors without medication and euthanasia, echo=TRUE}
Euthanasia_nomeds <- Euthanasia_donor %>% 
   filter(Antipsychotica_atypical == "NO" & Antidepressant_90days == "NO")
#number of donors 
length(unique(Euthanasia_nomeds$Donor_id))
#number of samples 
length(Euthanasia_nomeds$Donor_id)
```

