---
title: "Markers_all"
author: "LB"
date: "5/28/2021"
output: html_document
---

```{r load.packages, echo=FALSE, message=FALSE, results='hide'}
library(readxl)
library(edgeR)
library(limma)
library(DESeq2)
library(tidyverse)
library(data.table)
library(pheatmap)

knitr::opts_chunk$set( 
  warning=FALSE,
  message=FALSE,
  results = 'asis',
  error = FALSE,
  tidy = FALSE,
  fig.show = "hold")
```

```{r setup, echo=FALSE, message=FALSE, results='hide'}
#setwd("/Volumes/Samsung_T5/MiG_data/MDD_microglia_analysis/MDD_microglia_analysis/AD_analyses")
library(ProjectTemplate)
#load.project()
```

```{r directory}
data_dir <- "/Volumes/Samsung_T5/MiG_data/MDD_microglia_analysis/MDD_microglia_analysis/AD_analyses/data/"
project_dir <- "/Volumes/Samsung_T5/MiG_data/MDD_microglia_analysis/MDD_microglia_analysis/AD_analyses/"
```

```{r load_data, echo=TRUE, results = 'hide'}
load("/Volumes/Samsung_T5/MiG_data/MDD_microglia_analysis/MDD_microglia_analysis/AD_analyses/data/gene_matrix.RData")
metadata = "/Volumes/Samsung_T5/MiG_data/MDD_microglia_analysis/MDD_microglia_analysis/AD_analyses/data/metadata_gen_stats_alt.xslx"
metadata = read_excel(metadata, col_names = TRUE) 
metadata = as.data.frame(metadata)
#set rownames to donor_tissue
row.names(metadata) <- metadata$Donor_tissue
colnames(metadata)[16] <- "Medication"  

#this contains the samples removed after qc and lack of medication information 
load("/Volumes/Samsung_T5/MiG_data/MDD_microglia_analysis/MDD_microglia_analysis/AD_analyses/data/exclude_total.rda")



colnames(genes_tpm)[1] <- "13-072-CC" 
genes_tpm_filt = genes_tpm[, !colnames(genes_tpm) %in% exclude_total] 
metadata_filt = metadata[ !(rownames(metadata) %in% exclude_total), ]


#check for expression of microglia markers in dataset
markers_all = paste0(project_dir, "data/gene_markers.xlsx")
markers_all <- read_excel(markers_all, col_names = T)
markers_all = as.data.frame(markers_all)
colnames(markers_all) <- c("symbol", "category")

gencode_30 <- read.table("/Volumes/Samsung_T5/MiG_data/MDD_microglia_analysis/MDD_microglia_analysis/AD_analyses/data/ens.geneid.gencode.v30")
colnames(gencode_30) = c("ensembl","symbol")
gencode_30 <- gencode_30[-1,]


genes_tpm_filt = log2((genes_tpm_filt) + 1)
genes_tpm_filt <- as.data.frame(genes_tpm_filt)
setDT(genes_tpm_filt, keep.rownames = genes_tpm_filt$ensembl) 
#colnames(genes_tpm_filt), check if first col is called ensembl
names(genes_tpm_filt)[1] <- "ensembl"
res_name = merge(genes_tpm_filt, gencode_30, by="ensembl")
rownames(res_name) = res_name$ensembl

dim(genes_tpm_filt) #58929 239

#ensembl column is now redundant, so can be removed. 
genes_tpm_filt <- genes_tpm_filt[,-1]
dim(genes_tpm_filt) #58929 238
``` 

```{r data_load_markers, echo = TRUE, results = 'hide'}
marker_expression = merge(res_name, markers_all, by ="symbol")
dim (marker_expression) #293 240
head (marker_expression)
#check if category is included, if so, remove. 
metadata_filt_ordered <- metadata_filt[colnames(genes_tpm_filt),]
#metadata_filt_ordered = metadata_filt_ordered[-1,]
all(rownames(metadata_filt_ordered) == colnames (genes_tpm_filt)) #TRUE
Medication <- metadata_filt_ordered$Medication
```

### Boxplot IL10
```{r data_load_IL6, echo = TRUE}
IL10 <- marker_expression_inflammatory["IL10", ] 
IL10$ensembl = NULL
IL10$symbol = NULL
IL10_new<-as.data.frame(t(IL10))
dfIL10 = data.frame(Medication, IL10_new)
ggplot(data = dfIL10, mapping = aes(x = Medication, y = V1)) +
  geom_boxplot()+ 
  ggtitle("IL10") +
  stat_compare_means()+
  xlab("Medication") + ylab("log2((TPM)+1)")

```

### Boxplot IL1B
```{r data_load_IL1B, echo = TRUE}
IL1B <- marker_expression_inflammatory["IL1B", ] 
IL1B$ensembl = NULL
IL1B$symbol = NULL
IL1B_new<-as.data.frame(t(IL1B))
dfIL1B = data.frame(Medication, IL1B_new)
ggplot(data = dfIL1B, mapping = aes(x = Medication, y = V1)) +
  geom_boxplot()+ 
  stat_compare_means()+
  ggtitle("IL1B") +
  xlab("Medication") + ylab("log2((TPM)+1)")

```

``
### Boxplot IL6
```{r data_load_IL6, echo = TRUE}
IL6 <- marker_expression_inflammatory["IL6", ] 
IL6$ensembl = NULL
IL6$symbol = NULL
IL6_new<-as.data.frame(t(IL6))
dfIL6 = data.frame(Medication, IL6_new)
ggplot(data = dfIL6, mapping = aes(x = Medication, y = V1)) +
  geom_boxplot()+ 
  ggtitle("IL6") +
  stat_compare_means()+
  xlab("Medication") + ylab("log2((TPM)+1)")

```

### Boxplot NOS2
```{r data_load_NOS2, echo = TRUE}
NOS2 <- marker_expression_inflammatory["NOS2", ] 
NOS2$ensembl = NULL
NOS2$symbol = NULL
NOS2_new<-as.data.frame(t(NOS2))
dfNOS2 = data.frame(Medication, NOS2_new)
ggplot(data = dfNOS2, mapping = aes(x = Medication, y = V1)) +
  geom_boxplot()+ 
  ggtitle("NOS2") +
  stat_compare_means()+
  xlab("Medication") + ylab("log2((TPM)+1)")
  
 #add significance (ggpubber)

  

```

```{r data_load_TNF, echo = TRUE}
TNF <- marker_expression["TNF", ] 
TNF$ensembl = NULL
TNF$symbol = NULL
TNF_new<-as.data.frame(t(TNF))
dfTNF = data.frame(Medication, TNF_new)
ggplot(data = dfTNF, mapping = aes(x = Medication, y = V1)) +
  geom_boxplot()+ 
  ggtitle("TNF") +
  xlab("Medication") + ylab("log2((TPM)+1)")
```


# Heatmap with all markers  
### Both column and row clustering turned on

```{r data_load_heatmap_all, echo = TRUE, fig.width=14, fig.height=14, dpi=300}
marker_expression$ensembl = NULL
df_num = as.matrix(marker_expression[,2:239])
rownames(df_num) = sapply(marker_expression$symbol,function(x) 
strsplit(as.character(x),split = "\\\\")[[1]][1])
df_num_scale = scale(df_num)
df_num_scale = t(df_num_scale)
pos_df = data.frame("Medication" = metadata_filt_ordered$Medication)
rownames(pos_df) = rownames(df_num_scale)
pheatmap(df_num_scale, scale = "column", annotation_row = pos_df, main = "Target genes per medication", show_rownames = F)
```

### Row clustering 

```{r data_load_heatmap2_all, echo = TRUE, fig.width=14, fig.height=14, dpi=300}
pheatmap(df_num_scale, cluster_cols = F, cluster_rows = T, annotation_row = pos_df, main = "Target genes per medication", show_rownames = F)
```

### Row scaling 
#Scaling by row means that each row of the data matrix is taken in turn and given to the scale() function

```{r data_load_heatmap3_all, echo = TRUE, fig.width=14, fig.height=14, dpi=300}
pheatmap(df_num_scale, scale = "row", annotation_row = pos_df, main = "Target genes per stimulation", show_rownames = F)
```


Heatmap of only homeostatic genes 
```{r data_load_markers_homeostatic, echo = TRUE, results = 'hide'}
marker_homeostatic <- filter(markers_all, category == "Homeostatic")
#marker_homeostatic <- marker_homeostatic[,-2]

marker_expression_homeostatic = merge(res_name, marker_homeostatic, by ="symbol")
dim (marker_expression_homeostatic) #246 241
head (marker_expression_homeostatic)
#marker_expression_homeostatic <- marker_expression_homeostatic[, -#remove category column 
```
```{r}
#setdiff(marker_homeostatic$symbol, marker_expression_homeostatic$symbol)
# "FAM105A"  "HLA-DRB3" "HLA-DRB4" "TGFB1"  
# FAM105A > ENSG00000145569.6 voeg toe 
```


```{r data_load_heatmap, echo = TRUE, fig.width=14, fig.height=14, dpi=300}
marker_expression_homeostatic$ensembl = NULL
df_num = as.matrix(marker_expression_homeostatic[,2:239]) #c.1 is symbol, c.240 is category
rownames(df_num) = sapply(marker_expression_homeostatic$symbol,function(x) 
strsplit(as.character(x),split = "\\\\")[[1]][1])
df_num_scale = scale(df_num)
df_num_scale = t(df_num_scale)
pos_df = data.frame("Medication" = metadata_filt_ordered$Medication)
rownames(pos_df) = rownames(df_num_scale)
pheatmap(df_num_scale, scale = "column", annotation_row = pos_df, main = "Target homeostatic genes per medication", show_rownames = F)
```
### Row clustering 

```{r data_load_heatmap2, echo = TRUE, fig.width=14, fig.height=14, dpi=300}
pheatmap(df_num_scale, cluster_cols = F, cluster_rows = T, annotation_row = pos_df, main = "Target homeostatic genes per medication", show_rownames = F)
```
### Row scaling 
#Scaling by row means that each row of the data matrix is taken in turn and given to the scale() function

```{r data_load_heatmap3, echo = TRUE, fig.width=14, fig.height=14, dpi=300}
pheatmap(df_num_scale, scale = "row", annotation_row = pos_df, main = "Target homeostatic genes per stimulation", show_rownames = F)
```


Heatmap of only inflammatory genes 
```{r data_load_markers_inflammatory, echo = TRUE, results = 'hide'}
marker_inflammatory_p <- filter(markers_all, category == "Pro_inflammatory")
marker_inflammatory_a <- filter(markers_all, category == "Anti_inflammatory")
marker_inflammatory <- rbind(marker_inflammatory_p, marker_inflammatory_a) 


marker_expression_inflammatory = merge(res_name, marker_inflammatory, by ="symbol")
dim (marker_expression_inflammatory) #32 241
head (marker_expression_inflammatory)
marker_expression_inflammatory <- marker_expression_inflammatory[, -241]# remove category column 
```

```{r}
setdiff(marker_inflammatory$symbol, marker_expression_inflammatory$symbol)

```


```{r data_load_heatmap, echo = TRUE, fig.width=14, fig.height=14, dpi=300}
marker_expression_inflammatory$ensembl <- NULL
df_num <- as.matrix(marker_expression_inflammatory[,2:239]) #check voor invalid column (eg. category or sample)
rownames(df_num) <- sapply(marker_expression_inflammatory$symbol, function(x) 
strsplit(as.character(x),split = "\\\\")[[1]][1])
df_num_scale = scale(df_num)
df_num_scale = t(df_num_scale)
pos_df = data.frame("Medication" = metadata_filt_ordered$Medication)
rownames(pos_df) = rownames(df_num_scale)
pheatmap(df_num_scale, scale = "column", annotation_row = pos_df, main = "Target inflammatory genes per medication", show_rownames = F)
```
### Row clustering 

```{r data_load_heatmap2, echo = TRUE, fig.width=14, fig.height=14, dpi=300}
pheatmap(df_num_scale, cluster_cols = F, cluster_rows = T, annotation_row = pos_df, main = "Target inflammatory genes per medication", show_rownames = F)
```
### Row scaling 
#Scaling by row means that each row of the data matrix is taken in turn and given to the scale() function

```{r data_load_heatmap3, echo = TRUE, fig.width=14, fig.height=14, dpi=300}
pheatmap(df_num_scale, scale = "row", annotation_row = pos_df, main = "Target inflammatory genes per stimulation", show_rownames = F)
```

ALZHEIMER GENES 
Heatmap of only alzheimer genes 
```{r data_load_markers_alzheimer, echo = TRUE, results = 'hide'}
marker_alzheimer <- filter(markers_all, category == "Alzheimer")

marker_expression_alzheimer = merge(res_name, marker_alzheimer, by ="symbol")
dim (marker_expression_alzheimer) #46 241
head (marker_expression_alzheimer)
marker_expression_alzheimer <- marker_expression_alzheimer[, -241]# remove category column 
```

```{r}
#setdiff(marker_alzheimer$symbol, marker_expression_alzheimer$symbol)

```


```{r data_load_heatmap_alzheimer, echo = TRUE, fig.width=14, fig.height=14, dpi=300}
marker_expression_alzheimer$ensembl <- NULL
df_num_a <- as.matrix(marker_expression_alzheimer[,2:239]) #check voor invalid column (eg. category or sample)
rownames(df_num_a) <- sapply(marker_expression_alzheimer$symbol, function(x) 
strsplit(as.character(x),split = "\\\\")[[1]][1])
df_num_a_scale = scale(df_num_a)
df_num_a_scale = t(df_num_a_scale)
pos_df_a = data.frame("Medication" = metadata_filt_ordered$Medication)
rownames(pos_df_a) = rownames(df_num_a_scale)
pheatmap(df_num_a_scale, scale = "column", annotation_row = pos_df_a, main = "Target alzheimer genes per medication", show_rownames = F)
```
### Row clustering 

```{r data_load_heatmap2_alzheimer, echo = TRUE, fig.width=14, fig.height=14, dpi=300}
pheatmap(df_num_a_scale, cluster_cols = F, cluster_rows = T, annotation_row = pos_df_a, main = "Target alzheimer genes per medication", show_rownames = F)
```
### Row scaling 
#Scaling by row means that each row of the data matrix is taken in turn and given to the scale() function

```{r data_load_heatmap3_alzheimer, echo = TRUE, fig.width=14, fig.height=14, dpi=300}
pheatmap(df_num_a_scale, scale = "row", annotation_row = pos_df_a, main = "Target alzheimer genes per stimulation", show_rownames = F)
```