---
title: "PCA before filtering"
author: "LB"
date: "4/21/2021"
output: 
rmarkdown::html_document:
   theme: united
   highlight: tango
   code_folding: hide
   toc: true
   toc_float: true
   df_print: paged
   smooth_scroll: true
   number_sections: false
   self_contained: true
---

```{r load.packages, echo=FALSE, message=FALSE, results='hide'}
library(devtools)
library(factoextra)
library(ggfortify)
library(ggplot2)
library(dplyr)
library(broom)
library(variancePartition)
library(pheatmap)
library(edgeR)
library(limma)
library(DESeq2)
library(Hmisc)
library(matrixStats)
library(readxl)
library(tidyr)

knitr::opts_chunk$set( 
  warning=FALSE,
  message=FALSE,
  results = 'asis',
  error = FALSE,
  tidy = FALSE,
  fig.show = "hold")
```

```{r load_data, echo=TRUE, results = 'hide'}
load("/Volumes/Samsung_T5/MiG_data/MDD_microglia_analysis/MDD_microglia_analysis/AD_analyses/data/gene_matrix.RData")
metadata = "/Volumes/Samsung_T5/MiG_data/MDD_microglia_analysis/MDD_microglia_analysis/AD_analyses/data/metadata_gen_stats_alt.xslx"
metadata = read_excel(metadata, col_names = TRUE) 
metadata = as.data.frame(metadata)
#set rownames to sample 

row.names(metadata) <- metadata$Donor_tissue
#remove low count genes
cpm <- cpm(genes_counts) 
# CPM >= 1 in at least 50% of the samples
keep.exp <- rowSums(cpm > 1) >= (0.5 * ncol(genes_counts) )
genes_counts5 <- genes_counts[ keep.exp, ] #15575 genes 
#performing voom normalisation on data
counts_voom <- limma::voom(genes_counts5)
genes_counts_voom <- counts_voom$E
#order metadata and genes counts
#rownames(metadata)
#colnames(genes_counts_voom) 

#"13-072-CC" is the same sample as n"13-066-CC", so change name in genes_counts_voom
colnames(genes_counts_voom)[1] <- "13-072-CC"

# now we can order ! 

genes_counts_ordered <- genes_counts_voom[,rownames(metadata)]
#head(genes_counts_ordered)
all(rownames(metadata) == colnames (genes_counts_ordered)) #TRUE
```
```{r}
# save(genes_counts_ordered, file = "/Volumes/Samsung_T5/MiG_data/MDD_microglia_analysis/MDD_microglia_analysis/AD_analyses/data/genes_counts_ordered.Rdata")
```


```{r, echo=FALSE, results = 'hide'}
#in case we would like to to remove 13-066 > make new frames without this sample
#genes_counts_voom2 <- genes_counts_voom[,-1]
# same to remove 13-072 > 
#metadata_remove_13072 <- metadata[-1,] 

```

# PCAs
### PCA no scaling and centering of all 313 samples 

```{r PCA_noscale, echo=TRUE}
res.pca = prcomp(t(genes_counts_ordered))
autoplot(res.pca, colour = "Drugs", data = metadata)
```

### PCA with scaling and centering of all 313 samples 

```{r PCA_scale, echo=TRUE}
pca = prcomp(t(genes_counts_ordered), scale. = TRUE, center = TRUE)
autoplot(pca, colour = "Drugs", data = metadata)
```

### PCA with scaling and centering + labeling of all 313 samples

```{r PCA_labels, echo=TRUE}
pca = prcomp(t(genes_counts_ordered), scale. = TRUE, center = TRUE)
autoplot(pca, data= metadata, colour = 'Drugs', shape = FALSE, label.size = 2)
```

### Variances per PC of all samples 

```{r variances, echo=TRUE}
plot(pca)
```

# Boxplots of all samples
### PC1 

```{r boxplots_PC1, echo=TRUE}
PC1 <- pca$x[,1]
Drugs <- metadata$Drugs
df = data.frame(Drugs, PC1)
ggplot(data = df, mapping = aes(x = Drugs, y = PC1)) +
geom_boxplot()
```
### PC2  

```{r boxplots PC2, echo=TRUE}
PC2 <- pca$x[,2]
Drugs <- metadata$Drugs
df = data.frame(Drugs, PC2)
ggplot(data = df, mapping = aes(x = Drugs, y = PC2)) +
geom_boxplot()
```

# PCAs > removes NA
### PCA of samples with centering and scaling 

```{r PCA_cultured, echo=TRUE}
metadata_drugs <- metadata[!is.na(metadata$Drugs),] #50 samples
#check numbers
dim(metadata_drugs)
table(metadata_drugs$Drugs)
#check numbers per stimulation
table(metadata$Drugs)
#remove samples in genes counts datafile 
genes_counts_drugs <- genes_counts_ordered[,metadata_drugs$Donor_tissue]
pca = prcomp(t(genes_counts_drugs), scale. = TRUE, center = TRUE)
autoplot(pca, colour = "Drugs", data = metadata_drugs)
```

```{r, files overview, echo=FALSE, results='hide'}
# created metadata file without NA's about medication
# write.table(metadata_drugs, file = "/Volumes/Samsung_T5/MiG_data/MDD_microglia_analysis/MDD_microglia_analysis/AD_analyses/data/metadata_drugs_noNA.txt")

# I also made a file with the donor IDs that had no medication information > must be removed. 
# metadata_NA_drugs <- metadata[is.na(metadata$Drugs),]
#Donors_NA_drugs <- metadata_NA_drugs$Donor_tissue
# samples2remove_drug_NA <- unique(metadata_NA_drugs$Donor_tissue)
# write.table(samples2remove_drug_NA, file = "/Volumes/Samsung_T5/MiG_data/MDD_microglia_analysis/MDD_microglia_analysis/AD_analyses/data/samples2remove_drugNA.txt")

```

### PCA of samples with labeling of samples > NAs removed  

```{r PCA, labeling, echo=TRUE}
autoplot(pca, data = metadata_drugs, colour = 'Drugs', shape = FALSE, label.size = 2)
```

### Variances per PC of samples > NAs removed

```{r variances_no_nas, echo=TRUE}
plot(pca)
```

# Boxplots of cultured samples 
### PC1

```{r boxplots_1PC, echo=TRUE}
PC1 <- pca$x[,1]
Drugs <- metadata_drugs$Drugs
df = data.frame(Drugs, PC1)
ggplot(data = df, mapping = aes(x = Drugs, y = PC1)) +
  geom_boxplot()
```
### PC2 

```{r boxplots_PC2, echo=TRUE}
PC2 <- pca$x[,2]
Drugs <- metadata_drugs$Drugs
df = data.frame(Drugs, PC2)
ggplot(data = df, mapping = aes(x = Drugs, y = PC2)) +
  geom_boxplot()
```

# Scatterplot PC1 + PC2 

```{r scatterplot, echo=TRUE}
PCdf <- data.frame(sample = rownames(pca$x), pc1 = pca$x[,1], pc2 = pca$x[,2], Drugs = metadata_drugs$Drugs)
ggplot(data = PCdf, mapping = aes(x = PC1, y = PC2, colour = Drugs)) +
  geom_point()
```

#Heatmap with 49 PCs
```{r}
#set metadata to no capitals
names(metadata) = tolower(names(metadata))
indx <- sapply(metadata, is.character)
metadata[indx] <- lapply(metadata[indx], function(x) as.factor(x))

#include covariates
covariates = c( "donor_id",
                "diagnosis",
                "main_diagnosis",
                "tissue",
                "sex",
                "age",
                "pmd_minutes",
                "ph",
                "cause_of_death_categories",
                "smoking",
                "alcohol_dependence_daily_use",
                "autoimmune_diseases",
                "infection_2weeks",
                "drugs",
                "antidepressant_90days",
                "antidepressant_1",
                "antidepressant_2",
                "dose_mg", 
                "dose2_mg", 
                "antipsychotica_90days",
                "antipsychotica_typical",
                "antipsychotica_atypical",
                "dose_typical",
                "dose_atypical",
                "narcotics_90days",
                "benzo_90days",
                "other",
                "immuno",
                "ad last year?",
                "ad ever?",
                "picard_pct_mrna_bases",
                "picard_pct_ribosomal_bases",
                "picard_pct_pf_reads_aligned",
                "picard_summed_median",
                "picard_summed_mean")


#create format matrix 
matrix_rsquared = matrix(NA, nrow = length(covariates), ncol = 20) #Number of factors
matrix_pvalue = matrix(NA, nrow = length(covariates), ncol = 20)
#linear model function 
for (x in 1:length(covariates)){
  for (y in 1:20){
    matrix_rsquared[x,y] <- summary(lm(pca$x[,y] ~ metadata[,covariates[x]]) )$adj.r.squared
    matrix_pvalue[x,y] <- tidy(lm(pca$x[,y] ~ metadata[,covariates[x]]) )$p.value[2] #To insert pvalues in the heatmap
  }
}
#fill matrix with values 
matrix_rsquared <- as.data.frame(matrix_rsquared)
matrix_pvalue <- as.data.frame(matrix_pvalue)
rownames(matrix_rsquared) <- covariates
rownames(matrix_pvalue) <- covariates 
#create heatmap with rsquared values 
pheatmap(matrix_rsquared, main = "Correlation (Rsquared) between variables and first 20 PCs", legend = TRUE)
```
```
### Boxplot PC2 
#Variance mainly driven by ununstim

```{r boxplot_PC2, echo=TRUE}
PC2 <- pca$x[,2]
diagnosis <- metadata$diagnosis
df = data.frame(diagnosis, PC2)
ggplot(data = df, mapping = aes(x = diagnosis, y = PC2)) +
  geom_boxplot()
```

